---
title: "U.S. Border Crossings Analysis"
format: html
editor: visual
---

# Introduction hello hello hi hi hi my name ben 

**Core Research Question:**\
How have U.S. border crossings with Canada and Mexico evolved over time, and what do these patterns reveal about regional, economic, and policy-driven influences on cross-border mobility?

**Sub-Questions:** 1. How have crossing volumes changed from 1996 to the present across both borders? 2. Are there consistent seasonal or yearly trends? 3. Which states or specific ports account for the highest and lowest volumes? 4. Are certain regions more sensitive to changes than others, such as border tightening or travel bans? 5. How do crossing patterns differ by type (e.g., truck, rail, pedestrian, personal vehicle)? 6. What impact did specific nationwide or international events (e.g., 9/11, COVID-19, NAFTA revisions) have on crossing patterns? 7. Can we predict future crossing volumes based on historical trends or group ports/regions using clustering?

**START OF DATA MANAGEMENT AND FILE LOADING**

```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
border_data <- read_csv("Border.csv")
glimpse(border_data)

### Date clean
border_clean <- border_data %>%
  mutate(Date = my(Date))  # 'my' = month-year parser from lubridate

```

*Variable names and data management*

------------------------------------------------------------------------

```{r}
# RAW DATA: Read directly from the CSV
#border_data           # Original dataset from 'Border.csv'

######################################################################################################################

#  CLEANED DATA: Converted 'Date' to proper Date object (month-year)
#border_clean <- border_data %>%
  #mutate(Date = lubridate::my(Date))


######################################################################################################################
# (from border_clean):
# - Port Name     → Name of the border crossing port (chr)
# - State         → U.S. state where the port is located (chr)
# - Port Code     → Unique code assigned to the port (chr)
# - Border        → Either 'US-Canada Border' or 'US-Mexico Border' (chr)
# - Date          → Month-Year of the crossing (converted to Date)
# - Measure       → Type of crossing (e.g., Trucks, Pedestrians, Vehicles) (chr)
# - Value         → Number of crossings for that row (dbl)
# - Latitude      → GPS latitude of the port (dbl)
# - Longitude     → GPS longitude of the port (dbl)
# - Point         → Geometry point (chr; can be used for mapping)

# CLASSIFIED DATA: Tagged each crossing as 'Commercial' or 'Personal'
#border_classified <- border_clean %>%
  #mutate(TypeGroup = case_when(
    #str_detect(Measure, "Truck|Bus|Train|Rail") ~ "Commercial",
    #str_detect(Measure, "Pedestrian|Vehicle|Personal|Bicycl") ~ "Personal",
    #TRUE ~ "Other"
  #)) %>%
  #filter(TypeGroup != "Other")

######################################################################################################################

#Other possible helper columns:
# - month = lubridate::month(Date, label = TRUE)     # For seasonality
# - year  = lubridate::year(Date)                    # For trends or filters

#Tip: Use `glimpse(border_clean)` or `names(border_clean)` to re-check anytime

```

------------------------------------------------------------------------

## Data Managment

```{r}
library(qacData)
require(lubridate)
require(dplyr)
require(ggplot2)
require(stringr)

BorderCrossing <- read_csv("~/Documents/Exploratory Data Anaylsis/Border_Crossing_Entry_Data.csv")

BorderCrossing$Month <- substr(BorderCrossing$Date, 1,3)
BorderCrossing$Year <- substr(BorderCrossing$Date, 5,8)

BorderCrossing$Year <- as.numeric(BorderCrossing$Year)

BorderCrossing %>% 
  mutate(Monthn=case_when(Month=="Jan" ~ 1, Month=="Feb" ~2, Month=="Mar" ~3, Month=="Apr" ~4, Month=="May"~5, Month=="June" ~ 6, Month=="Jul" ~ 7, Month=="Aug" ~8, Month=="Sep" ~9, Month=="Oct" ~10, Month=="Nov"~11, Month=="Dec" ~ 12)) -> BorderCrossing


BorderCrossing %>% 
  mutate(Season=case_when(Monthn >2 & Monthn < 6 ~ "Spring", Monthn >5 & Month <9 ~ "Summer", Monthn >8 & Monthn <12 ~ "Fall", Monthn >11 | Monthn <3 ~ "Winter")) -> BorderCrossing

BorderCrossing %>%
  mutate(Date = make_date(Year, Monthn)) -> BorderCrossing

BorderCrossing %>% 
  mutate(format(Date, "%m-%Y")) -> BorderCrossing

names(BorderCrossing) <- c("PortName", "State", "PortCode","Border", "Dates", "Measure", "Value", "Latitude", "Longitude", "Point", "Months", "Year", "Month", "Season", "Date") 
```

# Jaime – Temporal Trends and Event Analysis

*Responsibilities:* - Analyze crossing volumes over time - Explore seasonal or yearly patterns - Investigate alignment with major events (9/11, COVID-19, etc.) - Use time series plots, trends, and event markers

```{r}

#Analysing crossing values over time

border_clean %>%
  group_by(Date, Border, Measure) %>%
  summarise(total = sum(Value), .groups = "drop")

######################################################################################################################


border_clean %>%
  group_by(Date, Border, Measure) %>%
  summarise(total = sum(Value), .groups = "drop") %>%
  ggplot(aes(x = Date, y = total, color = Measure)) +
  geom_line() +
  facet_wrap(~ Border, scales = "free_y") +
  labs(
    title = "Crossing Types Over Time by Border",
    x = "Date", y = "Total Crossings", color = "Type"
  ) +
  theme_minimal()

######################################################################################################################

```

```{r}
######################################################################################################################
#Exploring seasonal or yearly patterns
border_clean %>%
  mutate(month = month(Date, label = TRUE)) %>%
  group_by(month, Border) %>%
  summarise(avg_crossings = mean(Value, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = avg_crossings, fill = Border)) +
  geom_col(position = "dodge") +
  labs(title = "Average Crossings by Month (Seasonality)", y = "Avg Crossings") +
  theme_minimal()
######################################################################################################################
```

```{r}

######################################################################################################################

# Aggregate crossing totals by date and border
border_clean %>%
  group_by(Date, Border) %>%
  summarise(total = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Date, y = total, color = Border)) +
  geom_line(size = 1.2) +
  
  # highlighting major events in the U.S.
  geom_vline(xintercept = as.Date("2001-09-11"), linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "blue", size = 1) +
  
  
  annotate("text", x = as.Date("2001-09-11"), y = 15000000, label = "9/11", color = "red", angle = 90, vjust = -0.4, hjust = 0.5, fontface = "bold") +
  annotate("text", x = as.Date("2020-03-01"), y = 15000000, label = "COVID-19", color = "blue", angle = 90, vjust = -0.4, hjust = 0.5, fontface = "bold") +
  

  labs(
    title = "U.S. Border Crossings Over Time (with Major Events)",
    subtitle = "Dashed lines mark significant events: 9/11 and COVID-19",
    x = "Date",
    y = "Total Crossings",
    color = "Border"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 13)


######################################################################################################################
```

```{r}

######################################################################################################################
#See if commercial vehicles were imapcted or just pedestrians entering the U.S.
border_classified <- border_clean %>%
  mutate(
    TypeGroup = case_when(
      str_detect(Measure, "Truck|Bus|Train|Rail") ~ "Commercial",
      str_detect(Measure, "Pedestrian|Vehicle|Personal|Bicycl") ~ "Personal",
      TRUE ~ "Other"
    )
  ) %>%
  filter(TypeGroup != "Other")  # filter out junk or unclear categories

border_classified %>%
  group_by(Date, TypeGroup) %>%
  summarise(total = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Date, y = total, color = TypeGroup)) +
  geom_line(size = 1.2) +
  
  # Major event lines
  geom_vline(xintercept = as.Date("2001-09-11"), linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "blue", size = 1) +
  
  # Text labels for events
  annotate("text", x = as.Date("2001-09-11"), y = 15000000, label = "9/11", color = "red", angle = 90, vjust = -0.4, hjust = 0.5, fontface = "bold") +
  annotate("text", x = as.Date("2020-03-01"), y = 15000000, label = "COVID-19", color = "blue", angle = 90, vjust = -0.4, hjust = 0.5, fontface = "bold") +

  labs(
    title = "Border Crossings Over Time: Commercial vs Personal",
    subtitle = "Note steep declines aligned with major national events",
    x = "Date",
    y = "Total Crossings",
    color = "Crossing Type"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 13)

######################################################################################################################
```

# Ben Mayer – Regional and Port-Level Analysis

Responsibilities: hi my name is ben i made this code

Analyze crossing volume by state and port

Identify geographic variation and trends

Compare sensitivity to regional policies or restrictions

Use filtering, grouping, and plotting for geographic comparisons

```{r}

```

# Ben McKinney - Crossing Type and forecasting

Responsibilities:

Analyze patterns by crossing type

Identify how crossing types respond to events

(Optional) Try clustering ports based on traffic behavior

(Optional) Forecast future crossing volumes

Testing 123

```{r}
# Univariate Analysis: Percentage of total traffic volume by crossing type
univariate_data <- border_clean %>%
  group_by(Measure) %>%
  summarise(Total_Value = sum(Value, na.rm = TRUE)) %>%
  mutate(Percentage = Total_Value / sum(Total_Value) * 100)

# Bar plot for percentage analysis
ggplot(univariate_data, aes(x = Measure, y = Percentage, fill = Measure)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Percentage of Total Traffic Volume by Crossing Type", 
       x = "Crossing Type", y = "Percentage of Total Volume") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::percent_format(scale = 1))

# Statistical Test: ANOVA to determine if there are significant differences in traffic volume by crossing type
anova_result <- aov(Value ~ Measure, data = border_clean)
summary(anova_result)

```

```{r}
# Filter for one year before and after March 2020
border_periods <- border_clean %>%
  filter(Date >= as.Date("2019-03-01") & Date < as.Date("2021-03-01")) %>%
  mutate(Period = case_when(
    Date < as.Date("2020-03-01") ~ "Year Before Mar 2020",
    TRUE ~ "Year After Mar 2020"
  ))

# Summarize average monthly crossings by Measure and Period
average_summary <- border_periods %>%
  group_by(Measure, Period) %>%
  summarise(Average_Monthly = mean(Value, na.rm = TRUE), .groups = 'drop')

# Plot comparison of average monthly crossings
comparison_plot <- ggplot(average_summary, aes(x = Measure, y = Average_Monthly, fill = Period)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Average Monthly Crossings by Type (Year Before vs. After March 2020)",
       x = "Crossing Type",
       y = "Average Monthly Crossings") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Print the comparison plot
print(comparison_plot)

# Prepare monthly totals per period
monthly_totals <- border_periods %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Measure, Period, Month) %>%
  summarise(Monthly_Total = sum(Value, na.rm = TRUE), .groups = 'drop')

# Get unique Measures
measures <- unique(monthly_totals$Measure)
```

```{r}
# Run t-tests for each Measure
test_results <- lapply(measures, function(m) {
  test_data <- monthly_totals %>% filter(Measure == m)
  t_out <- t.test(Monthly_Total ~ Period, data = test_data)
  data.frame(
    Measure = m,
    Before_Mean = t_out$estimate[1],
    After_Mean = t_out$estimate[2],
    P_Value = t_out$p.value
  )
})

# Combine results into a single data frame
test_summary <- do.call(rbind, test_results)

# Print the test summary
print(test_summary)

# Grab significant results
significant_measures <- test_summary %>%
  filter(P_Value < 0.05) %>%
  pull(Measure)

# Print only the statistically significant comparisons
print(significant_measures)

```

# Conclusion

Summarize major findings from each section, compare insights across regions/types/time, and reflect on how this connects to broader patterns in mobility and policy.
