---
title: "U.S. Border Crossings Analysis"
format: html
editor: visual
---

# Introduction

**Core Research Question:**\
How have U.S. border crossings with Canada and Mexico evolved over time, and what do these patterns reveal about regional, economic, and policy-driven influences on cross-border mobility?

**Sub-Questions:** 1. How have crossing volumes changed from 1996 to the present across both borders? 2. Are there consistent seasonal or yearly trends? 3. Which states or specific ports account for the highest and lowest volumes? 4. Are certain regions more sensitive to changes than others, such as border tightening or travel bans? 5. How do crossing patterns differ by type (e.g., truck, rail, pedestrian, personal vehicle)? 6. What impact did specific nationwide or international events (e.g., 9/11, COVID-19, NAFTA revisions) have on crossing patterns? 7. Can we predict future crossing volumes based on historical trends or group ports/regions using clustering?

**START OF DATA MANAGEMENT AND FILE LOADING**

*Variable names and data management*

```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(qacData)
library(knitr)
require(lubridate)
require(stringr)

# ========================
# RAW DATA
# ========================
border_data <- read_csv("Border.csv")

# ============================================================
# CLEANED DATA: Convert 'Date', extract helpers, classify time (Ben why did you do this so manually)
# ============================================================
border_clean <- border_data %>%
  # Convert raw 'Date' like "Jan-1996" to Date object
  mutate(
    Date = lubridate::my(Date),
    Month = substr(Date, 1, 3),                   
    Year = lubridate::year(Date),                 
    Monthn = lubridate::month(Date),              
    MonthFormatted = format(Date, "%m-%Y"),       
    Season = case_when(                            
      Monthn >= 3 & Monthn <= 5 ~ "Spring",
      Monthn >= 6 & Monthn <= 8 ~ "Summer",
      Monthn >= 9 & Monthn <= 11 ~ "Fall",
      Monthn == 12 | Monthn <= 2 ~ "Winter",
      TRUE ~ "Unknown"
    )
  )

# ==================================================================================
# CLASSIFIED DATA: Group crossings into 'Commercial' and 'Personal' traffic types
# ==================================================================================
border_classified <- border_clean %>%
  mutate(TypeGroup = case_when(
    str_detect(Measure, "Truck|Bus|Train|Rail") ~ "Commercial",
    str_detect(Measure, "Pedestrian|Vehicle|Personal|Bicycl") ~ "Personal",
    TRUE ~ "Other"
  )) %>%
  filter(TypeGroup != "Other")

```

------------------------------------------------------------------------

```{r}
# RAW DATA: Read directly from the CSV
#border_data           # Original dataset from 'Border.csv'

######################################################################################################################

#  CLEANED DATA: Converted 'Date' to proper Date object (month-year)
#border_clean <- border_data %>%
  #mutate(Date = lubridate::my(Date))


######################################################################################################################
# (from border_clean):
# - Port Name     → Name of the border crossing port (chr)
# - State         → U.S. state where the port is located (chr)
# - Port Code     → Unique code assigned to the port (chr)
# - Border        → Either 'US-Canada Border' or 'US-Mexico Border' (chr)
# - Date          → Month-Year of the crossing (converted to Date)
# - Measure       → Type of crossing (e.g., Trucks, Pedestrians, Vehicles) (chr)
# - Value         → Number of crossings for that row (dbl)
# - Latitude      → GPS latitude of the port (dbl)
# - Longitude     → GPS longitude of the port (dbl)
# - Point         → Geometry point (chr; can be used for mapping)

# CLASSIFIED DATA: Tagged each crossing as 'Commercial' or 'Personal'
#border_classified <- border_clean %>%
  #mutate(TypeGroup = case_when(
    #str_detect(Measure, "Truck|Bus|Train|Rail") ~ "Commercial",
    #str_detect(Measure, "Pedestrian|Vehicle|Personal|Bicycl") ~ "Personal",
    #TRUE ~ "Other"
  #)) %>%
  #filter(TypeGroup != "Other")

######################################################################################################################

#Other possible helper columns:
# - month = lubridate::month(Date, label = TRUE)     # For seasonality
# - year  = lubridate::year(Date)                    # For trends or filters

#Tip: Use `glimpse(border_clean)` or `names(border_clean)` to re-check anytime

```

------------------------------------------------------------------------

## Data Managment

# Jaime – Temporal Trends and Event Analysis

*Responsibilities:* - Analyze crossing volumes over time - Explore seasonal or yearly patterns - Investigate alignment with major events (9/11, COVID-19, etc.) - Use time series plots, trends, and event markers

```{r}

library(scales)

border_clean %>%
  group_by(Date, Border, Measure) %>%
  summarise(total = sum(Value), .groups = "drop") %>%
  filter(Measure %in% c("Pedestrians", "Personal Vehicle Passengers", "Personal Vehicles", "Truck Containers Loaded", "Trucks")) %>%
  ggplot(aes(x = Date, y = total, color = Measure)) +
  geom_line(size = 1) +
  facet_wrap(~ Border, scales = "free_y") +
  scale_y_continuous(labels = label_comma()) +
  scale_color_manual(values = c(
    "Pedestrians" = "#1b9e77",
    "Personal Vehicle Passengers" = "#d95f02",
    "Personal Vehicles" = "#7570b3",
    "Truck Containers Loaded" = "#e7298a",
    "Trucks" = "#66a61e"
  )) +
  labs(
    title = "Top 5 Border Crossing Types Over Time",
    x = "Date", y = "Total Crossings", color = "Type"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")


######################################################################################################################

```

This graph shows the top five types of border crossings over time at the U.S.-Canada and U.S.-Mexico borders. **Personal vehicle passengers and personal vehicles consistently represent the highest crossing volumes** at both borders, with sharp seasonal patterns especially evident at the northern border. Overall, the **U.S.-Mexico border consistently sees higher crossing volumes**, especially for personal vehicle passengers, while pedestrian and commercial crossings like truck container traffic remain lower but relatively stable over time.

```{r}
library(ggplot2)
library(dplyr)
library(scales)

# Group by Season and Border
seasonal_data <- border_clean %>%
  group_by(Season, Border) %>%
  summarise(avg_crossings = mean(Value, na.rm = TRUE), .groups = "drop")

#logic
seasonal_data$Season <- factor(seasonal_data$Season, levels = c("Winter", "Spring", "Summer", "Fall"))


ggplot(seasonal_data, aes(x = Season, y = avg_crossings, fill = Border)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Border, scales = "free_y") +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Average Border Crossings by Season",
    x = "Season",
    y = "Average Crossings"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(size = 12, face = "bold"))


```

This visualization compares the average number of border crossings by season at the U.S.-Canada and U.S.-Mexico borders. **Crossings at the U.S.-Canada border show a distinct summer peak**, reflecting seasonal travel and tourism patterns, while winter months experience the lowest volumes. In contrast, **the U.S.-Mexico border shows minimal seasonal variation**, maintaining high crossing levels across all seasons. This suggests that **cross-border activity at the southern border is more consistent year-round**, likely due to ongoing economic, social, and commuting needs.

```{r}
library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)

# monthly averages by each border
monthly_data <- border_clean %>%
  mutate(Month = month(Date, label = TRUE, abbr = TRUE)) %>%
  group_by(Month, Border) %>%
  summarise(avg_crossings = mean(Value, na.rm = TRUE), .groups = "drop")

# month order logic
monthly_data$Month <- factor(monthly_data$Month, levels = month.abb)


ggplot(monthly_data, aes(x = Month, y = avg_crossings, fill = Border)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ Border, scales = "free_y") +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Average Border Crossings by Month",
    x = "Month",
    y = "Average Crossings"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(size = 12, face = "bold")
  )

```

These two plots show the average number of border crossings per month for the U.S.-Canada and U.S.-Mexico borders. At the **U.S.-Canada border**, there is a clear seasonal pattern, with crossings peaking in the **summer months (June–August)**—likely due to tourism and favorable weather—and dropping during the winter. In contrast, the **U.S.-Mexico border maintains high crossing volumes year-round**, with only minor fluctuations across months. This suggests that **economic and social factors**, rather than seasonality, drive most of the traffic at the southern border.

```{r}
######################################################################################################################

# Aggregate crossing totals by date and border
border_clean %>%
  group_by(Date, Border) %>%
  summarise(total = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Date, y = total, color = Border)) +
  
  # Add main trend lines
  geom_line(size = 1.2) +

  # Optional: Add smoothed trend overlay for visibility
  geom_smooth(se = FALSE, method = "loess", span = 0.15, linetype = "dotted", size = 0.6, alpha = 0.6) +

  # Add shaded region for extended COVID-era restrictions
  annotate("rect", xmin = as.Date("2020-03-01"), xmax = as.Date("2021-06-01"),
           ymin = -Inf, ymax = Inf, fill = "blue", alpha = 0.05) +

  # Vertical lines for major events
  geom_vline(xintercept = as.Date("2001-09-11"), linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "blue", size = 1) +

  # Annotate events with text
  annotate("text", x = as.Date("2001-09-11"), y = 15000000,
           label = "9/11", color = "red", angle = 90, vjust = -0.4,
           hjust = 0.5, fontface = "bold") +
  annotate("text", x = as.Date("2020-03-01"), y = 15000000,
           label = "COVID-19", color = "blue", angle = 90, vjust = -0.4,
           hjust = 0.5, fontface = "bold") +

  # Labels and styling
  labs(
    title = "U.S. Border Crossings Over Time (with Major Events)",
    subtitle = "Dashed lines mark significant events: 9/11 and COVID-19",
    caption = "Source: U.S. Border Crossing Data (BTS)",
    x = "Date",
    y = "Total Crossings",
    color = "Border"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 13)

######################################################################################################################

```

This time-series visualization highlights U.S. border crossing trends at the northern and southern borders from the late 1990s to the present, with major national events marked for context. The **red dashed line** indicates the 9/11 attacks (2001), after which **a noticeable and sustained decline in crossings at the U.S.-Canada border** is observed. In contrast, the U.S.-Mexico border, while also affected, rebounded more quickly. The **blue dashed line and shaded region** represent the onset and extended disruption caused by **COVID-19 in early 2020**, which resulted in an immediate and historic drop in crossings at both borders. Recovery began around mid-2021 but has not yet fully returned to pre-pandemic levels. Overall, the **U.S.-Mexico border consistently experiences higher crossing volumes**, and the chart shows how both borders respond differently to external shocks, likely due to differences in travel purpose, policy enforcement, and economic dependency.

```{r}

######################################################################################################################

# Classify crossings as Commercial vs Personal
border_classified <- border_clean %>%
  mutate(
    TypeGroup = case_when(
      str_detect(Measure, "Truck|Bus|Train|Rail") ~ "Commercial",
      str_detect(Measure, "Pedestrian|Vehicle|Personal|Bicycl") ~ "Personal",
      TRUE ~ "Other"
    )
  ) %>%
  filter(TypeGroup != "Other")  # filter out unclear or rare categories

# Plot total crossings over time by TypeGroup
border_classified %>%
  group_by(Date, TypeGroup) %>%
  summarise(total = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Date, y = total, color = TypeGroup)) +
  
  # Main lines
  geom_line(size = 1.2) +

  # Optional smoothed trend (comment out if you want cleaner lines only)
  geom_smooth(se = FALSE, method = "loess", span = 0.2, linetype = "dotted", alpha = 0.3) +

  # Shaded region for extended COVID impact
  annotate("rect", xmin = as.Date("2020-03-01"), xmax = as.Date("2021-06-01"),
           ymin = -Inf, ymax = Inf, alpha = 0.05, fill = "blue") +

  # Major event vertical lines
  geom_vline(xintercept = as.Date("2001-09-11"), linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "blue", size = 1) +

  # Event labels
  annotate("text", x = as.Date("2001-09-11"), y = 15000000, label = "9/11",
           color = "red", angle = 90, vjust = -0.4, hjust = 0.5, fontface = "bold") +
  annotate("text", x = as.Date("2020-03-01"), y = 15000000, label = "COVID-19",
           color = "blue", angle = 90, vjust = -0.4, hjust = 0.5, fontface = "bold") +

  # Titles and styling
  labs(
    title = "Border Crossings Over Time: Commercial vs Personal",
    subtitle = "Note steep declines aligned with major national events",
    caption = "Source: U.S. Border Crossing Data (BTS)",
    x = "Date",
    y = "Total Crossings",
    color = "Crossing Type"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 13)

######################################################################################################################

```

This graph compares commercial and personal border crossings over time, highlighting their distinct responses to major national events. **Personal crossings show sharp, immediate declines following both 9/11 and the onset of COVID-19**, reflecting heightened security measures and travel restrictions that primarily impacted non-essential travel. **Commercial crossings, however, remained relatively stable throughout**, even during the peak of the COVID-19 pandemic, underscoring their essential role in maintaining supply chains and economic continuity. The post-2020 recovery in personal crossings is evident but gradual, while commercial traffic exhibits remarkable consistency across decades, largely insulated from external shocks.

# Ben Mayer – Regional and Port-Level Analysis

Responsibilities:

Analyze crossing volume by state and port

Identify geographic variation and trends

Compare sensitivity to regional policies or restrictions

Use filtering, grouping, and plotting for geographic comparisons

```{r}

library(ggplot2)
library(dplyr)
library(maps)
library(scales)

# cleaning state data
BorderCrossing1 <- border_clean %>%
  group_by(State) %>%
  summarise(TotalCrossings = sum(Value, na.rm = TRUE)) %>%
  mutate(State = tolower(State))

us_states <- map_data("state")

map_data_joined <- left_join(us_states, BorderCrossing1, by = c("region" = "State"))


ggplot(map_data_joined, aes(long, lat, group = group, fill = TotalCrossings)) +
  geom_polygon(color = "black") +
  scale_fill_gradient(
    low = "#ffffcc", high = "#ff4444", na.value = "gray90",
    labels = label_comma(),  # 👈 NO SCIENTIFIC NOTATION
    name = "Total Crossings"
  ) +
  labs(
    title = "Total U.S. Border Crossings by State",
    subtitle = "Based on land ports of entry at U.S.-Mexico and U.S.-Canada borders",
    caption = "Source: U.S. Border Crossing Data (BTS)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold"),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

  
```

This graph shows the number of border crossings across the different states. This indicates that the data set has more crossings on the Mexico border than the Canada border. Also the Canada border crossing are spread out while the Mexico crossings are constricted to just 4 states.

```{r}
library(ggplot2)
library(dplyr)
library(maps)
library(scales)
library(patchwork) 

#base map
us_states <- map_data("state")

# border per region
border_data_grouped <- border_clean %>%
  filter(Border %in% c("US-Mexico Border", "US-Canada Border")) %>%
  group_by(State, Border) %>%
  summarise(TotalCrossings = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  mutate(State = tolower(State))

# map base
make_map <- function(border_name, title_label, fill_palette) {
  border_states <- border_data_grouped %>% filter(Border == border_name)
  map_join <- left_join(us_states, border_states, by = c("region" = "State"))

  ggplot(map_join, aes(long, lat, group = group)) +
    geom_polygon(aes(fill = TotalCrossings), color = "black") +
    scale_fill_gradient(
      low = "#ffffcc", high = fill_palette, na.value = "gray90",
      labels = label_comma(), name = paste0("Total Crossings\n(", border_name, ")")
    ) +
    labs(title = title_label) +
    coord_fixed(1.3) +
    theme_minimal(base_size = 13) +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold"),
      legend.title = element_text(size = 10),
      legend.position = "right"
    )
}


map_canada <- make_map("US-Canada Border", "U.S.-Canada Border", "#2b8cbe")
map_mexico <- make_map("US-Mexico Border", "U.S.-Mexico Border", "#de2d26")

#patchwork logic
map_canada + map_mexico + plot_annotation(
  title = "Total Border Crossings by State (Separate Color Scales)",
  subtitle = "Each map uses its own scale and legend for comparison clarity",
  caption = "Source: U.S. Border Crossing Data (BTS)"
)


```

These two choropleth maps show the total number of border crossings by state, separated by the **U.S.-Mexico** and **U.S.-Canada** borders, with each map using its own color scale for clarity. On the **southern border**, **Texas and California dominate** in total crossings, reflecting their large, busy ports of entry and proximity to major cities and trade routes. In contrast, on the **northern border**, states like **New York, Michigan, and Washington** show the highest volumes, driven by both commercial transport and tourism. Splitting the maps this way highlights the **different geographic patterns and scales** of activity at each border, and avoids the distortion that occurs when all states are placed on a single color scale.

```{r}
library(ggplot2)
library(dplyr)
library(scales)

# Prepare data
border_data <- border_clean %>%
  group_by(Border) %>%
  summarise(TotalCrossings = sum(Value, na.rm = TRUE)) %>%
  mutate(
    Percentage = TotalCrossings / sum(TotalCrossings),
    Label = case_when(
      Border == "US-Mexico Border" ~ "🇲🇽 Mexico Border",
      Border == "US-Canada Border" ~ "🇨🇦 Canada Border"
    ),
    LabelText = paste0(round(Percentage * 100), "%\n", comma(TotalCrossings))
  )

# Donut chart with numbers and percentages
ggplot(border_data, aes(x = 2, y = TotalCrossings, fill = Label)) +
  geom_col(width = 1, color = "black") +
  coord_polar("y", start = 0) +
  xlim(0.5, 2.5) +  # Creates the donut hole
  scale_fill_manual(values = c("🇲🇽 Mexico Border" = "#f16913", "🇨🇦 Canada Border" = "#3182bd")) +
  geom_text(
    aes(label = LabelText),
    position = position_stack(vjust = 0.5),
    color = "black", size = 5, fontface = "bold", lineheight = 1.2
  ) +
  labs(
    title = "Share of Total U.S. Land Border Crossings",
    subtitle = "With percentage and total crossings labeled",
    fill = "Border",
    caption = "Source: U.S. Border Crossing Data (BTS)"
  ) +
  theme_void(base_size = 15) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.title = element_text(face = "bold"),
    legend.position = "right"
  )

```

This graph shows the relationship between number of border crossings and which border they crossed. The Mexico border has over twice as many crossings then the Canada border. I wonder if including flying border crossings would change the results of this distribution.

```{r}

library(ggplot2)
library(dplyr)
library(scales)
library(gganimate)


BorderCrossing3 <- border_clean %>%
  group_by(Year, State) %>%
  summarise(TotalCrossings = sum(Value), .groups = "drop") %>%
  mutate(Year = as.numeric(Year))

#filter top 8 states for clarity
top_states <- BorderCrossing3 %>%
  group_by(State) %>%
  summarise(total = sum(TotalCrossings)) %>%
  top_n(8, total) %>%
  pull(State)

BorderCrossing3 <- filter(BorderCrossing3, State %in% top_states)


animated_plot <- ggplot(BorderCrossing3, aes(x = Year, y = TotalCrossings, color = State, group = State)) +
  geom_line(linewidth = 1.3) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Animated Border Crossings by State",
    subtitle = "Total crossings per year (Top States)",
    x = "Year", y = "Total Crossings", color = "State"
  ) +
  theme_minimal(base_size = 14) +
  transition_reveal(Year)

animate(animated_plot, width = 800, height = 500, fps = 10, renderer = gifski_renderer())



```

This graph shows the relationship between years and the border crossings by state. The top throughout are Texas and California. Every state sees dips around 2008 and 2020. There is also a dip during 2025 but that's because the data in 2025 is incomplete. The top 2 states experience bigger drops then the lower 6 states. The states at the top 2 states are Mexico.

```{r}
ggplot(BorderCrossing3, aes(x = Year, y = TotalCrossings, fill = State)) +
  geom_area(position = "stack", alpha = 0.9) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Total Border Crossings Over Time by State",
    subtitle = "Stacked to show relative contributions per year",
    x = "Year",
    y = "Total Crossings",
    fill = "State"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )

```

```{r}
library(ggplot2)
library(dplyr)
library(scales)
library(viridis)


BorderCrossing3 <- border_clean %>%
  group_by(Year, State) %>%
  summarise(TotalCrossings = sum(Value), .groups = "drop") %>%
  mutate(Year = as.numeric(Year))

#  top 10 states by total crossings
top_states <- BorderCrossing3 %>%
  group_by(State) %>%
  summarise(total = sum(TotalCrossings)) %>%
  top_n(10, total) %>%
  pull(State)

BorderCrossing3_filtered <- BorderCrossing3 %>%
  filter(State %in% top_states)


ggplot(BorderCrossing3_filtered, aes(x = Year, y = reorder(State, -TotalCrossings), fill = TotalCrossings)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(label = comma, option = "C") +
  geom_vline(xintercept = 2001, linetype = "dashed", color = "red", size = 0.4) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "blue", size = 0.4) +
  labs(
    title = "Heatmap of Border Crossings by State and Year",
    subtitle = "Top 10 border states • Dotted lines mark 9/11 and COVID-19",
    x = "Year",
    y = "State",
    fill = "Total Crossings"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    plot.subtitle = element_text(hjust = 0.5),
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```

```{r}
library(ggplot2)
library(dplyr)
library(treemapify)
library(scales)


BorderCrossing4 <- border_clean %>%
  group_by(State, Measure) %>%
  summarise(TotalCrossings = sum(Value), .groups = "drop")

#top states only
top_states <- BorderCrossing4 %>%
  group_by(State) %>%
  summarise(StateTotal = sum(TotalCrossings)) %>%
  top_n(6, StateTotal) %>%
  pull(State)

BorderCrossing_treemap <- BorderCrossing4 %>%
  filter(State %in% top_states)


ggplot(BorderCrossing_treemap, aes(
  area = TotalCrossings,
  fill = Measure,
  label = paste(Measure, "\n", comma(TotalCrossings)),
  subgroup = State
)) +
  geom_treemap() +
  geom_treemap_subgroup_border(color = "white") +
  geom_treemap_subgroup_text(place = "top", grow = TRUE, alpha = 0.6, color = "black", fontface = "bold") +
  geom_treemap_text(colour = "white", place = "centre", reflow = TRUE, size = 12/.pt) +
  labs(
    title = "Border Crossing Types by State",
    subtitle = "Treemap grouped by State (top 6 states by total crossings)",
    fill = "Crossing Type"
  ) +
  theme_minimal(base_size = 13)


```

This graph shows the relationship between the top 6 state and border crossing type. In every state the most popular is personal vehicles and there passengers. Pedestrians is the 3rd most popular in the Mexico border states but isn't very present in the Canada border states. Trucks is instead the 3rd most popular for the Canada border states but isn't very present in the Mexico Border States.4

## Ben McKinney - Crossing Type and forecasting

Responsibilities:

Analyze patterns by crossing type

Identify how crossing types respond to events

(Optional) Try clustering ports based on traffic behavior

(Optional) Forecast future crossing volumes

```{r}
# Univariate Analysis: Percentage of total traffic volume by crossing type
univariate_data <- border_clean %>%
  group_by(Measure) %>%
  summarise(Total_Value = sum(Value, na.rm = TRUE)) %>%
  mutate(Percentage = Total_Value / sum(Total_Value) * 100)

kable(univariate_data, digits = 2, 
      caption = "Percentage of Total Traffic Volume by Crossing Type")

# Bar plot for percentage analysis
ggplot(univariate_data, aes(x = Measure, y = Percentage, fill = Measure)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Percentage of Total Traffic Volume by Crossing Type", 
       x = "Crossing Type", y = "Percentage of Total Volume") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(labels = scales::percent_format(scale = 1))

# Statistical Test: ANOVA to determine if there are significant differences in traffic volume by crossing type
anova_result <- aov(Value ~ Measure, data = border_clean)
summary(anova_result)

```

The table and bar chart above are a basic univariate analysis representing the number of total crossings by type. Overall, Personal vehicles and Personal Vehicle Passengers accounts for 80% of the total crossings for both borders. Commercial transport, such as rail/truck containers and trucks, accounts for 6.37% of the total border crossings.

```{r}
# Filter for one year before and after March 2020
border_periods <- border_clean %>%
  filter(Date >= as.Date("2019-03-01") & Date < as.Date("2021-03-01")) %>%
  mutate(Period = case_when(
    Date < as.Date("2020-03-01") ~ "Year Before Mar 2020",
    TRUE ~ "Year After Mar 2020"
  ))

# Summarize average monthly crossings by Measure and Period
average_summary <- border_periods %>%
  group_by(Measure, Period) %>%
  summarise(Average_Monthly = mean(Value, na.rm = TRUE), .groups = 'drop')

# Plot comparison of average monthly crossings
comparison_plot <- ggplot(average_summary, aes(x = Measure, y = Average_Monthly, fill = Period)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  labs(title = "Average Monthly Crossings by Type (Year Before vs. After March 2020)",
       x = "Crossing Type",
       y = "Average Monthly Crossings") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Print the comparison plot
print(comparison_plot)

# Prepare monthly totals per period
monthly_totals <- border_periods %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Measure, Period, Month) %>%
  summarise(Monthly_Total = sum(Value, na.rm = TRUE), .groups = 'drop')

# Get unique Measures
measures <- unique(monthly_totals$Measure)
```

The bar chart above compares the average number of of border crossings by type before and after the covid pandemic began in March of 2020. Overall, nearly every crossing type decreased drastically after the pandemic began. The most drastic of these involved pedestrians and personal vehicles/passengers. Conducting a T-test, we can see that 10 out of 12 crossing types had statistically significant decreases in volume. The only 2 exceptions to this pattern are truck containers emptied and trucks, which did not significantly change after Covid. This is likely due to the necessity of supply chain goods coming into the US. Additionally, border restrictions may not have affected all types of freight equally, which may be evidenced by the significant decrease in rail containers.

```{r}
# Run t-tests for each Measure
test_results <- lapply(measures, function(m) {
  test_data <- monthly_totals %>% filter(Measure == m)
  t_out <- t.test(Monthly_Total ~ Period, data = test_data)
  data.frame(
    Measure = m,
    Before_Mean = t_out$estimate[1],
    After_Mean = t_out$estimate[2],
    P_Value = t_out$p.value
  )
})

# Combine results into a single data frame
test_summary <- do.call(rbind, test_results)

# Print the test summary
print(test_summary)

# Grab significant results
significant_measures <- test_summary %>%
  filter(P_Value < 0.05) %>%
  pull(Measure)

# Print only the statistically significant comparisons
print(significant_measures)

```

# Conclusion

Summarize major findings from each section, compare insights across regions/types/time, and reflect on how this connects to broader patterns in mobility and policy.
