---
title: "U.S. Border Crossings Analysis"
format: html
editor: visual
---

# Introduction

**Core Research Question:**\
How have U.S. border crossings with Canada and Mexico evolved over time, and what do these patterns reveal about regional, economic, and policy-driven influences on cross-border mobility?

**Sub-Questions:** 1. How have crossing volumes changed from 1996 to the present across both borders? 2. Are there consistent seasonal or yearly trends? 3. Which states or specific ports account for the highest and lowest volumes? 4. Are certain regions more sensitive to changes than others, such as border tightening or travel bans? 5. How do crossing patterns differ by type (e.g., truck, rail, pedestrian, personal vehicle)? 6. What impact did specific nationwide or international events (e.g., 9/11, COVID-19, NAFTA revisions) have on crossing patterns? 7. Can we predict future crossing volumes based on historical trends or group ports/regions using clustering?

**START OF DATA MANAGEMENT AND FILE LOADING**

```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
border_data <- read_csv("Border.csv")
glimpse(border_data)

### Date clean
border_clean <- border_data %>%
  mutate(Date = my(Date))  # 'my' = month-year parser from lubridate

```

------------------------------------------------------------------------

# Jaime – Temporal Trends and Event Analysis

*Responsibilities:* - Analyze crossing volumes over time - Explore seasonal or yearly patterns - Investigate alignment with major events (9/11, COVID-19, etc.) - Use time series plots, trends, and event markers

```{r}

#Analysing crossing values over time

border_clean %>%
  group_by(Date, Border, Measure) %>%
  summarise(total = sum(Value), .groups = "drop")

######################################################################################################################


border_clean %>%
  group_by(Date, Border, Measure) %>%
  summarise(total = sum(Value), .groups = "drop") %>%
  ggplot(aes(x = Date, y = total, color = Measure)) +
  geom_line() +
  facet_wrap(~ Border, scales = "free_y") +
  labs(
    title = "Crossing Types Over Time by Border",
    x = "Date", y = "Total Crossings", color = "Type"
  ) +
  theme_minimal()

######################################################################################################################

```
```{r}
######################################################################################################################
#Exploring seasonal or yearly patterns
border_clean %>%
  mutate(month = month(Date, label = TRUE)) %>%
  group_by(month, Border) %>%
  summarise(avg_crossings = mean(Value, na.rm = TRUE)) %>%
  ggplot(aes(x = month, y = avg_crossings, fill = Border)) +
  geom_col(position = "dodge") +
  labs(title = "Average Crossings by Month (Seasonality)", y = "Avg Crossings") +
  theme_minimal()
######################################################################################################################
```

```{r}

######################################################################################################################

# Aggregate crossing totals by date and border
border_clean %>%
  group_by(Date, Border) %>%
  summarise(total = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Date, y = total, color = Border)) +
  geom_line(size = 1.2) +
  
  # highlighting major events in the U.S.
  geom_vline(xintercept = as.Date("2001-09-11"), linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "blue", size = 1) +
  
  
  annotate("text", x = as.Date("2001-09-11"), y = 15000000, label = "9/11", color = "red", angle = 90, vjust = -0.4, hjust = 0.5, fontface = "bold") +
  annotate("text", x = as.Date("2020-03-01"), y = 15000000, label = "COVID-19", color = "blue", angle = 90, vjust = -0.4, hjust = 0.5, fontface = "bold") +
  

  labs(
    title = "U.S. Border Crossings Over Time (with Major Events)",
    subtitle = "Dashed lines mark significant events: 9/11 and COVID-19",
    x = "Date",
    y = "Total Crossings",
    color = "Border"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 13)


######################################################################################################################
```

```{r}

######################################################################################################################
#See if commercial vehicles were imapcted or just pedestrians entering the U.S.
border_classified <- border_clean %>%
  mutate(
    TypeGroup = case_when(
      str_detect(Measure, "Truck|Bus|Train|Rail") ~ "Commercial",
      str_detect(Measure, "Pedestrian|Vehicle|Personal|Bicycl") ~ "Personal",
      TRUE ~ "Other"
    )
  ) %>%
  filter(TypeGroup != "Other")  # filter out junk or unclear categories

border_classified %>%
  group_by(Date, TypeGroup) %>%
  summarise(total = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Date, y = total, color = TypeGroup)) +
  geom_line(size = 1.2) +
  
  # Major event lines
  geom_vline(xintercept = as.Date("2001-09-11"), linetype = "dashed", color = "red", size = 1) +
  geom_vline(xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "blue", size = 1) +
  
  # Text labels for events
  annotate("text", x = as.Date("2001-09-11"), y = 15000000, label = "9/11", color = "red", angle = 90, vjust = -0.4, hjust = 0.5, fontface = "bold") +
  annotate("text", x = as.Date("2020-03-01"), y = 15000000, label = "COVID-19", color = "blue", angle = 90, vjust = -0.4, hjust = 0.5, fontface = "bold") +

  labs(
    title = "Border Crossings Over Time: Commercial vs Personal",
    subtitle = "Note steep declines aligned with major national events",
    x = "Date",
    y = "Total Crossings",
    color = "Crossing Type"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 13)

######################################################################################################################
```



# Ben Mayer – Regional and Port-Level Analysis

Responsibilities:

Analyze crossing volume by state and port

Identify geographic variation and trends

Compare sensitivity to regional policies or restrictions

Use filtering, grouping, and plotting for geographic comparisons

```{r}

```

# Ben McKinney - Crossing Type and forecasting

Responsibilities:

Analyze patterns by crossing type

Identify how crossing types respond to events

(Optional) Try clustering ports based on traffic behavior

(Optional) Forecast future crossing volumes

Testing 123

```{r}

```

# Conclusion

Summarize major findings from each section, compare insights across regions/types/time, and reflect on how this connects to broader patterns in mobility and policy.
